---
title: "Non-violent emergency call outcome analysis in Eugene, Oregon"
subtitle: "Investigating Eugene's response to addiction, homelessness and mental health related crisis"
author: "Mathilde Lafay"
date: "`r Sys.Date()`"
output:
  pdf_document:
    toc: true
    toc_depth: 3
    number_sections: true
    fig_caption: true
    df_print: kable
    latex_engine: xelatex
    pandoc_args: ["--syntax-highlighting", "tango"]
include-before:
  - \newpage
---
\newpage

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = FALSE,
  warning = FALSE,
  message = FALSE
)
```

```{r libraries}

library(here)
library(tidyverse)
library(tidytext)
library(dplyr)
library(scales)
library(readxl)
library(purrr)
library(janitor)
library(lubridate)
library(hms)
library(patchwork)

library(fixest)   
library(did)   
library(stargazer)
library(broom)

```

```{r data}
data <- readRDS(here("data", "clean", "data_merged.rds"))
```

```{r color palette}

# color blind friendly palette https://colors-tester.com/free/fr/
color_agencies <- c(
  "EPD"     = "#2ACCFE",  
  "SPD"     = "#FEA4E6",  
  "CAHOOTS" = "#58BF03",  
  "MCSLC"   = "#FF8C66"   
)

```

# Introduction 

In this document, we will be conducting a quantitative analysis of the call outcomes for mental health, addiction and homelessness related emergencies in Eugene. We will be focusing on the code and methods. For more details about the project, please find the full report at the root of the repository: `internship_report.pdf`.

# Work Environment

## Prerequisites
- [R](https://cran.r-project.org/) (version 4.5 or higher recommended)
- [RStudio](https://posit.co/download/rstudio-desktop/) (optional, but recommended)

## Managing Dependencies
This project uses [`renv`](https://rstudio.github.io/renv/) for R package management. To install all required packages:

1. Open the project in RStudio or start an R session in the project directory.
2. Run:
	```R
	renv::restore()
	```
	This will install all packages listed in the project's lockfile.

If you add new packages, use `renv::snapshot()` to update the lockfile.

# Data

The data used in this analysis is processed and merged from the following sources. The final dataset includes every agency (EPD, SPD, CAHOOTS and MCSLC), and dispatched calls only, from 2015-01-01 to 2025-12-31.

## Available raw data

- SPD Call for Services, 2015-2025 : .xlsx document, one sheet per year
- SPD Responding Units, 2015-2025 : .xlsx document, one sheet per year
- SPD Call Outcome, 2015-2025 : .xlsx doxument, key codes and one sheet per year

- MCS LC data, 2024/08/18 - 2025/12/17 : .xlsx document, one sheet

- Eugene CAD data, 2015-2025 : one .csv per year

### SPD data

This dataset contains detailed records of calls for service for Springfield Police Department from 2015 to 2025, including temporal management of calls, the nature of the interventions, and the mobile resources deployed.

There are 14 columns and 567,655 rows.

| Variable               | Type      | Example Value         | Description                                                                    |
|------------------------|-----------|-----------------------|--------------------------------------------------------------------------------|
| `timestamp`            | datetime  | 2015-01-01 01:16:21   | Date and time when the incident was initially recorded.                        |
| `incident_id`          | double    | 15000074              | Unique identifier for each incident.                                           |
| `agency`               | factor    | SPD                   | Agency responsible for the intervention.                                       |
| `city`                 | factor    | Springfield           | City.                                                                          |
| `nature_code`          | factor    | ALMAUD                | Coded category for the type of call received.                                  |
| `nature_final`         | factor    | AUDIBLE ALARM         | Final classification of the incident after assessment.                         |
| `close_code`           | factor    | BLDS                  | Coded reason or outcome for closing the incident.                              |
| `close_code_definition`| factor    | Building Check Secure | Full description of the incident closure status.                               |
| `priority`             | factor    | 3                     | Numerical priority level assigned to the call.                                 |
| `prime_unit`           | character | 1S18                  | Primary unit assigned to respond to the incident.                              |
| `units`                | character | 1S18, 1S21, K94       | List of all units dispatched to the scene.                                     |
| `dispatch_time`        | time      | 01:18:00              | Time at which units were dispatched.                                           |
| `arrival_time`         | time      | 01:21:00              | Time at which the first unit arrived at the scene.                             |
| `clear_time`           | time      | 01:32:00              | Time at which the incident was marked as completed.                            |
| `call_source`          | factor    | PHONE                 | Origin or method of the incoming call.                                         |



### Eugene CAD data

This dataset contains detailed records of calls for service for Eugene from 2015 to 2025, including temporal management of calls, the nature of the interventions, and the mobile resources deployed.

There are 17 columns and 1,446,014 rows.

| Variable              | Type      | Example Value         | Description                                                                    |
|-----------------------|-----------|-----------------------|--------------------------------------------------------------------------------|
| `timestamp`           | datetime  | 2015-01-01 00:00:00   | Date and time when the call to report an incident was initially recorded.      |
| `incident_id`         | double    | 15000001              | Unique identifier for each incident.                                           |
| `agency`              | factor    | EPD                   | Agency responsible for the intervention (e.g., Eugene Police Department).      |
| `service`             | factor    | LAW                   | Type of service mobilized (e.g., Law Enforcement, Fire, Other).                |
| `call_source`         | factor    | SELF                  | Origin of the call (e.g., Self-initiated, Phone, W911).                        |
| `nature`              | factor    | PERSON STOP           | Nature or reported motive of the incident.                                     |
| `close_code`          | factor    | ASST                  | Abbreviated closing code indicating the outcome of the intervention.           |
| `closed_as`           | factor    | ASSISTED              | Full description of how the incident was resolved.                             |
| `priority`            | factor    | 6                     | Priority level assigned to the incident ranging from 1 (highest) to 9 (lowest).|
| `dispatch_time`       | time      | 00:00:00              | Time when units were dispatched to the incident location.                      |
| `arrival_time`        | time      | 00:00:00              | Time when the first unit arrived on the scene.                                 |
| `clear_time`          | time      | 00:03:37              | Time when the incident was marked as cleared or completed.                     |
| `dispatch`            | factor    | 1                     | Binary indicator specifying whether a unit was dispatched.                     |
| `arrival`             | factor    | 1                     | Binary indicator specifying whether a unit arrived on the scene.               |
| `prime_unit`          | factor    | _5E48                 | Identifier of the primary unit assigned to the incident.                       | 
| `nb_units_dispatched` | double    | 1                     | Total number of units dispatched to the intervention.                          |
| `nb_units_arrived`    | double    | 1                     | Total number of units that successfully reached the scene.                     |


### Mobile Crisis Services - Lane County data

This dataset tracks crisis intervention activities from MCS LC, focusing on response timelines, and clinical outcomes for mobile crisis units.

| Variable                     | Type      | Example Value         | Description                                                                 |
|------------------------------|-----------|-----------------------|-----------------------------------------------------------------------------|
| `timestamp`                  | datetime  | 2025-12-17 02:57:00   | Initial date and time the crisis call was logged.                           |
| `incident_id`                | double    | 13209                 | Unique identifier for the crisis incident.                                  |
| `agency`                     | factor    | MCS LC                | The specific Mobile Crisis Service agency responding to the call.           |
| `city`                       | factor    | Springfield           | The city or municipality where the incident occurred.                       |
| `call_type`                  | factor    | Agitation             | The primary clinical nature or behavioral reason for the call.              |
| `call_outcome`               | factor    | Remained in community | The final disposition or result of the intervention.                        |
| `dispatch_time`              | datetime  | 2025-12-17 03:14:00   | Time when the mobile crisis team was officially dispatched.                 |
| `client_engagement_time`     | datetime  | 2025-12-17 03:38:00   | Time when the team made direct clinical contact with the client.            |
| `arrival_time`               | datetime  | 2025-12-17 03:35:00   | Time when the mobile crisis unit arrived at the scene.                      |
| `clear_time`                 | datetime  | 2025-12-17 03:48:00   | Time when the unit completed the intervention and became available again.   |
| `dispatch_status`            | factor    | Engaged               | Status of the dispatch (e.g., Engaged, Refused).                            |
| `minutes_request_dispatch`   | double    | 17                    | Duration in minutes between the initial request and dispatch.               |
| `minutes_dispatch_arrival`   | double    | 21                    | Duration in minutes between dispatch and arrival on scene.                  |
| `minutes_arrival_engagement` | double    | 3                     | Duration in minutes between arrival and start of client engagement.     |
| `minutes_arrival_departure`  | double    | 13                    | Total time spent on scene in minutes (arrival to departure).                |


# Overview and exploratory analysis

The following analysis will focus on investigating the changes in Eugene's response to non-violent emergencies related to homelessness, addiction and mental health after MCS LC's launch on August 18th 2024 and CAHOOTS' discontinuation on April 7th 2025 in Eugene.

## Summary

Let's begin with a generic overview and exploratory analysis of the data.

```{r glimpse}
glimpse(data)
```

## Global Call Volume

```{r global call volume}

data %>%
  group_by(city) %>%
  summarise(
    Total_calls = n(),
    EPD   = sum(EPD == 1),
    SPD   = sum(SPD == 1),
    CAHOOTS = sum(CAHOOTS == 1),
    MCSLC = sum(MCSLC == 1)
  ) %>%
  knitr::kable(caption = "Global repartition of call volume per city and agency")
```

## Frequent Natures

The most frequent natures for CAHOOTS calls are Public Assist and Check Welfare. These will be our main focus for the whole analysis, along with some other call natures appearing in the plot. Total, we will use the top 14 most frequent natures in calls handled by CAHOOTS in both Eugene and Springfield, and the mcslc natures (14).

```{r cahoots natures, fig.align = "center", fig.width = 7, fig.height = 4.5, fig.cap = "Top 10 most frequent call natures for CAHOOTS dispatches"}

data_nature_cahoots <- data %>%
  filter(CAHOOTS == 1) %>%
  count(nature, sort = TRUE) %>%
  top_n(20, n)

ggplot(data_nature_cahoots, aes(x = reorder(nature, n), y = n)) +
  geom_col(fill = "#58BF03", width = 0.7) +
  geom_text(
    aes(label = scales::comma(n)),
    hjust = 1.1,                   
    color = "#1D3557",             
    size = 3,                     
    fontface = "bold"
  ) +
  coord_flip() +
  labs(
    title = "Most frequent natures (CAHOOTS)",
    x = "Nature of incident",
    y = "Number of interventions"
  )

```


```{r most frequent natures}

top_14_cahoots_natures <- c(
  "Public Assist",
  "Check Welfare",
  "Transport",
  "Suicidal Subject",
  "Intoxicated Subject",
  "Disorderly Subject",
  "Found Syringe",
  "Info / Atl",
  "Disoriented Subject",
  "Dispute",
  "Subject Down",
  "Disorderly Juveniles",
  "Suspicious Conditions",
  "Police Officer Hold"
)

top_mcslc_natures <- c(
  "Agitation Or Disruptive Behavior",
  "Disorganized Behavior",
  "Difficulty Functioning",
  "Needing Social/Mental Health Services",
  "Suicidal Subject",
  "Harm/Risk Of Harm To Self/Others/Property",
  "Adult Social/Interpersonal Problems",
  "Paranoia",
  "Substance Use",
  "Other",
  "Concerns About Treatment Engagement",
  "Trauma",
  "Running Away",
  "Seeking Mental Health Services"
)

```

## Evolution In Time

Here is a timeline of CAHOOTS' and MCSLC's dispatched calls evolution between 2015 and 2025. We can see that right after CAHOOTS stopped, MCSLC's number of dispatches went up.

```{r evolution in time,fig.align = "center", fig.width = 7.5, fig.height = 4.5, fig.cap = "Monthly evolution of call volume by mcs agency (2015-2025)"}

remarkable_dates <- as.Date(c("2024-08-18", "2025-04-07"))

data_chronologie <- data %>%
  filter(city == "Eugene") %>%
  mutate(Mois = floor_date(timestamp, "month")) %>%
  group_by(Mois) %>%
  summarise(
    CAHOOTS = sum(CAHOOTS == 1),
    MCS_LC  = sum(MCSLC == 1)
  ) %>%
  pivot_longer(cols = c(CAHOOTS, MCS_LC), names_to = "Service", values_to = "Volume")

ggplot(data_chronologie, aes(x = Mois, y = Volume, color = Service)) +
  geom_vline(xintercept = as.POSIXct(remarkable_dates), 
           linetype = "dashed", color = "red", size = 0.7) +
  geom_text(data = data.frame(dates = remarkable_dates, label = c("MCSLC", " CAHOOTS")),
            aes(x = dates, y = max(data_chronologie$Volume), label = label),
            inherit.aes = FALSE, angle = 90, vjust = -0.5, color = "red") +
  coord_cartesian(clip = "off") +
  geom_line(size = 0.8) +
  scale_color_manual(values = color_agencies) +  
  labs(
    title = "Timeline for Mobile Crisis Services",
    subtitle = "Monthly intervention volume",
    x = "Chronology (Month)",
    y = "Number of dispatched calls"
  ) +
  theme_minimal() +
  theme(plot.margin = unit(c(1, 1, 1, 1), "cm"))

```

## Broad Categories Overview
Next is a general overview of broad outcome categories for every existing call nature, for the whole dataset (EPD, SPD, MCSLC and CAHOOTS responses).  

```{r plot aggregated outcomes, fig.width = 10, fig.height = 5.5}

# COLOR PALETTE
color_outcomes <- c(
  "Arrest"               = "#E63946", 
  "Emergency Department" = "#F4A261",  
  "Community Support"    = "#2A9D8F",  
  "Other"                = "#A8DADC"
)

# CREATE AGGREGATED OUTCOMES
data_outcomes <- data %>%
  mutate(
    has_police = if_else(EPD == 1 | SPD == 1, TRUE, FALSE),
    has_crisis = if_else(CAHOOTS == 1 | MCSLC == 1, TRUE, FALSE),
    response_profile = case_when(
      has_police & has_crisis ~ "Joint Response",
      has_police & !has_crisis ~ "Police Only",
      !has_police & has_crisis ~ "Crisis Unit Only",
      TRUE ~ "Other / Missing"
    )
  ) %>%
  filter(response_profile != "Other / Missing") %>%
  mutate(
    outcome_clean = str_to_lower(outcome),
    macro_outcome = case_when(
      str_detect(outcome_clean, "arrest|custody|citation") ~ "Arrest",
      str_detect(outcome_clean, "transport|hospital|medical|department|detox") ~ "Emergency Department",
      str_detect(outcome_clean, "assist|counseling|resolved|community|advised|welfare") ~ "Community Support",
      TRUE ~ "Other"
    ),
    macro_outcome = factor(macro_outcome, levels = c("Arrest", "Emergency Department", "Community Support", "Other"))
  ) %>%
  group_by(response_profile, macro_outcome) %>%
  summarise(Count = n(), .groups = "drop") %>%
  group_by(response_profile) %>%
  mutate(Proportion = Count / sum(Count))

# PLOT STACKED BARS
ggplot(data_outcomes, aes(x = response_profile, y = Proportion, fill = macro_outcome)) +
  geom_col(position = "fill", width = 0.6) +
  scale_y_continuous(labels = percent_format()) +
  scale_fill_manual(values = color_outcomes) +
  coord_flip() + 
  labs(
    title = "Call Outcomes Distribution by Response Profile",
    subtitle = "Relative proportion of final dispositions (100% stacked)",
    x = "Response Profile",
    y = "Percentage of Total Calls",
    fill = "Outcome"
  ) +
  theme_minimal(base_size = 12) +
  theme(
    plot.title = element_text(face = "bold", size = 14, color = "#1D3557"),
    plot.subtitle = element_text(size = 11, color = "#4A4A4A"),
    panel_arrest.background = element_rect(fill = "#FFFFFF", color = NA),
    plot.background = element_rect(fill = "#FFFFFF", color = NA),
    panel_arrest.grid.major.y = element_blank(), 
    panel_arrest.grid.minor = element_blank(),
    legend.position = "top",        
    axis.text.y = element_text(face = "bold", color = "#1D3557")
  )
```


## Exploratory Analysis of Call Outcome 

To better understand the effects of CAHOOTS' discontinuation and MCSLC's roll-out, we will be focusing on the time period where they were both active, and then the period during which CAHOOTS stopped servicing Eugene. This will allow us to see if there is any change in MCSLC's activity, if other agencies, like the EPD, took over some calls that CAHOOTS used to handle, and how it impacted call outcomes in general.

### Temporal Filtering and Harmonization

First, let's filter the data to keep only the relevant time periods: 
- CAHOOTS and MCSLC active: August 18, 2024 to April 7, 2025
- After CAHOOTS stopped services: April 8, 2025 to present

To do so, we define the temporal boundaries in datetime variables. We then group the data into weekly bins to smooth out daily fluctuations, and we sort the the calls into the `Overlap` and the `Post-CAHOOTS` categories, respectively the MCSLC and CAHOOTS joint period, and the only MCSLC period. We exclude all historical data, before August 18, 2024 when MCSLC started to maintain a strictly controlled analytical scope.


```{r temporal filtering}
date_mcslc_start <- as.POSIXct("2024-08-18")
date_cahoots_stop <- as.POSIXct("2025-04-07")
date_end <- as.POSIXct("2025-12-17") 

data_temporal <- data %>%
  mutate(
    week = floor_date(timestamp, "week"),
    period = case_when(
      timestamp >= date_mcslc_start & timestamp <= date_cahoots_stop ~ "Overlap",
      timestamp > date_cahoots_stop & timestamp <= date_end ~ "Post-CAHOOTS",
      TRUE ~ "Exclude"
    )
  ) %>%
  filter(period != "Exclude")

```
The overlap period is 232 days long, and the post-CAHOOTS period has 254 days.

```{r days count, include = FALSE}

date_mcslc_start <- as.POSIXct("2024-08-18")
date_cahoots_stop <- as.POSIXct("2025-04-07")
date_end <- as.POSIXct("2025-12-17")

days_overlap <- as.numeric(difftime(date_cahoots_stop, date_mcslc_start, units = "days"))

days_post <- as.numeric(difftime(date_end, date_cahoots_stop, units = "days"))
```


We have 9038 dispatches for CAHOOTS and 1116 dispatches for MCSLC during the overlap period. During the post-CAHOOTS period, we have 5894 MCSLC dispatches.

```{r count dispatches per period, include = FALSE}
data_temporal %>%
  filter(period == "Overlap", city == "Eugene", CAHOOTS == 1) %>%
  count()
data_temporal %>%
  filter(period == "Overlap", city == "Eugene", MCSLC == 1) %>%
  count()
data_temporal %>%
  filter(period == "Post-CAHOOTS", city == "Eugene", MCSLC == 1) %>%
  count()
```

Here is an overview of the filtered data.

For MCLSC dispatches, we have the 7 following possible outcomes. Most of dispatched calls end with **Remaine In Community**, and we note that some calls end with **Arrest**.

```{r MCSLC temporal outcomes}
data_temporal %>%
  filter(MCSLC == 1) %>%
  count(outcome) %>%
  slice_max(n, n = 7)
```

CAHOOTS only calls have 35 possible outcomes. Here are the 10 most frequent ones, with **Assisted** and **Unable To Locate being the most frequent ones**. For CAHOOTS, most of the interventions are closed as **Assisted** with no further detail. Some outcomes are from CAHOOTS and EPD joint responses. EPD interventions are often closed with different outcomes for the same type of call, we can assume that these outcomes are similar but just registered under a different name.

```{r CAHOOTS temporal outcomes}
data_temporal %>%
  filter(CAHOOTS == 1, EPD == 0) %>%
  count(outcome) %>%
  slice_max(n, n = 36)
```


### CAHOOTS vs MCSLC

#### Overlap Period Comparison

The goal here is to look at CAHOOTS and MCSLC's types of services. We extract the most frequent call natures for both agencies. For this time period and city, CAHOOTS has 12 call natures, with **Public Assist** and **Check Welfare** being the most common ones. MCSLC has 14, **Agitation or Disruptive Behavior** and **Disorganized Behavior** are the two most common.

```{r top 14 natures}
data_nature_overlap <- data_temporal %>%
  filter(period == "Overlap", city == "Eugene") %>%
  filter(CAHOOTS == 1 | MCSLC == 1) %>%
  mutate(agency = case_when(
    CAHOOTS == 1 & MCSLC == 0 ~ "CAHOOTS",
    MCSLC == 1 & CAHOOTS == 0 ~ "MCSLC",
  )) %>%
  filter(!is.na(agency)) %>%
  count(agency, nature, sort = TRUE) %>%
  group_by(agency) %>%
  slice_max(n, n = 20) %>%
  mutate(nature = reorder_within(nature, n, agency))


ggplot(data_nature_overlap, aes(x = nature, y = n, fill = agency)) +
  geom_col(show.legend = FALSE) +
  scale_x_reordered() +
  scale_fill_manual(values = color_agencies) +
  facet_wrap(~agency, scales = "free_y") +
  coord_flip() +
  labs(
    title = "Most frequent call natures by agency during overlap period",
    subtitle = "Aug 18, 2024 – Apr 7, 2025 | Eugene",
    x = NULL, y = "Number of dispatched calls"
  ) +
  theme_minimal()
```


Here is the call outcome distribution by agency during the overlap period.

```{r}
data_outcome_overlap <- data_temporal %>%
  filter(period == "Overlap", city == "Eugene") %>%
  filter(CAHOOTS == 1 | MCSLC == 1) %>%
  mutate(agency = case_when(
    CAHOOTS == 1 & MCSLC == 0 ~ "CAHOOTS",
    MCSLC == 1 & CAHOOTS == 0 ~ "MCSLC",
    CAHOOTS == 1 & MCSLC == 1 ~ "Joint"
  )) %>%
  filter(!is.na(agency)) %>%
  count(agency, outcome) %>%
  group_by(agency) %>%
  mutate(prop = n / sum(n)) %>%
  group_by(agency) %>%
  slice_max(n, n = 10)

ggplot(data_outcome_overlap, aes(x = reorder_within(outcome, prop, agency), 
                                  y = prop, fill = agency)) +
  geom_col(show.legend = FALSE) +
  scale_x_reordered() +
  scale_y_continuous(labels = percent_format(accuracy = 1)) +
  scale_fill_manual(values = color_agencies) +
  facet_wrap(~agency, scales = "free_y") +
  coord_flip() +
  labs(
    title = "Outcome distribution by agency during overlap period",
    subtitle = "Aug 18, 2024 – Apr 7, 2025 | Eugene",
    x = NULL, y = "Share of calls"
  ) +
  theme_minimal()
```

If we reuse the broad outcome category from the exploratory analysis, we find the following numbers. We will dive in deeper and more precisely in the following subsections of this analysis.

```{r}
data_temporal %>%
  filter(period == "Overlap", city == "Eugene") %>%
  filter(CAHOOTS == 1 | MCSLC == 1) %>%
  mutate(
    agency = case_when(
      CAHOOTS == 1 & MCSLC == 0 ~ "CAHOOTS",
      MCSLC == 1 & CAHOOTS == 0 ~ "MCSLC",
      TRUE ~ "Joint"
    ),
    outcome_clean = str_to_lower(as.character(outcome)),
    macro_outcome = case_when(
    str_detect(outcome_clean, "arrest|custody|citation") ~ "Arrest",
    
    str_detect(outcome_clean, "emergency department|hospital|medical|detox|sobering") ~ "ED / Crisis Facility",
    outcome_clean == "crisis walk-in center" ~ "ED / Crisis Facility",
    outcome_clean == "respite" ~ "ED / Crisis Facility",
    
    str_detect(outcome_clean, "assist|advised|resolved|counseling|welfare check done") ~ "Community Support",
    outcome_clean == "remained in community" ~ "Community Support",
    
    str_detect(outcome_clean, "unable to locate|gone on arrival|disregard") ~ "No Contact",
    
    str_detect(outcome_clean, "refused") ~ "Refused Services",
    
    TRUE ~ "Other"
  )
  ) %>%
  count(agency, macro_outcome) %>%
  group_by(agency) %>%
  mutate(prop = scales::percent(n / sum(n), accuracy = 0.1)) %>%
  arrange(agency, desc(n)) %>%
  knitr::kable(
    caption = "Broad outcome category distribution by agency during overlap period (Eugene, Aug 2024 – Apr 2025)",
    col.names = c("Agency", "Outcome", "Count", "Share")
  )
```

#### Breakdown of Call Outcome


```{r outcome distribution plot}
make_outcome_nature_dual <- function(df, agency_col, agency_label, top_n_natures) {
  
  top_natures <- df %>%
    filter(period == "Overlap", city == "Eugene", .data[[agency_col]] == 1) %>%
    count(nature, sort = TRUE) %>%
    slice_head(n = top_n_natures) %>%
    pull(nature)
  
  df_plot <- df %>%
    filter(period == "Overlap", city == "Eugene", .data[[agency_col]] == 1,
           nature %in% top_natures) %>%
    count(nature, outcome) %>%
    group_by(nature) %>%
    mutate(
      prop = n / sum(n),
      nature = factor(nature, levels = rev(top_natures))
    )
  
  p_abs <- ggplot(df_plot, aes(x = nature, y = n, fill = outcome)) +
    geom_col(position = "stack", width = 0.7) +
    scale_y_continuous(labels = comma) +
    coord_flip() +
    labs(
      title = "Absolute counts",
      x = NULL, y = "Number of calls", fill = "Outcome"
    ) +
    theme_minimal(base_size = 10) +
    theme(
      legend.position = "none",
      plot.title = element_text(face = "bold", size = 10)
    )
  
  p_pct <- ggplot(df_plot, aes(x = nature, y = prop, fill = outcome)) +
    geom_col(position = "fill", width = 0.7) +
    scale_y_continuous(labels = percent_format(accuracy = 1)) +
    coord_flip() +
    labs(title = "Relative proportions", x = NULL, y = "Share of calls", fill = "Outcome") +
    theme_minimal(base_size = 10) +
    theme(
      legend.position = "bottom",
      legend.direction = "horizontal",    # Forcer l'horizontal
      legend.box = "horizontal",          # Modifier ceci
      legend.text = element_text(size = 6), # Réduire un peu plus
      legend.title = element_blank(),     # Enlever le titre pour gagner de la place
      legend.key.size = unit(0.2, "cm"),
      plot.title = element_text(face = "bold", size = 10),
      axis.text.y = element_blank()
    )
  
  p_abs + p_pct +
    plot_layout(widths = c(2.5, 1), guides = "collect") +
    plot_annotation(
      title = paste("Call outcome breakdown by nature —", agency_label),
      subtitle = "Overlap period: Aug 18, 2024 – Apr 7, 2025 | Eugene only",
      theme = theme(
        legend.position = "bottom",
        plot.title = element_text(face = "bold", size = 13),
        plot.subtitle = element_text(size = 10, color = "#4A4A4A")
      )
    )
}

```


During the overlap period, CAHOOTS handled a majority of **Public Assist** and **Check Welfare** calls, representing over half of all CAHOOTS calls in Eugene for this time period. The dominant outcome across almost all natures is **Assisted**, confirming that most CAHOOTS interventions resulted in direct on-scene support without escalation. 

However, notable exceptions appear for **Info/Atl** and **Disoriented Subject**, where outcomes like **Gone On Arrival** and **Unable To Locate** are more frequent. **Arrest** is rare across all natures. 



```{r outcomes distribution CAHOOTS}
plot_cahoots_outcomes <- make_outcome_nature_dual(
  data_temporal, 
  "CAHOOTS", 
  "CAHOOTS", 
  top_n_natures = 12)

plot_cahoots_outcomes
```
MCSLC's call volume during the overlap period is lower than CAHOOTS'. It could be due to its more recent launch and its separate dispatch pathway (they are outside of the 911 CAD system). While MCSLC's goal is to manage mental health, behavioral, and substance abuse crisis as well, their scope of action stays different than CAHOOTS'.

The dominant outcome across all natures is **Remained in Community**. **Emergency Department** transport appears as a secondary outcome, especially for higher-risk natures such as **Suicidal Subect** and **Harm/Risk of Harm** calls.

We also see some **Arrest** outcomes for some call natures, indicating a joint response from MCSLC and the EPD, even though they do not appear as responding agency in the data. MCSLC being on its own call line is not jointly dispatched along with the police, but they coordinate with other agencies if needed (EPD, other existing care providers...).


```{r outcomes distribution MCSLC}
plot_mcslc_outcomes <- make_outcome_nature_dual(
  data_temporal, 
  "MCSLC", 
  "MCSLC", 
  top_n_natures = 14)

plot_mcslc_outcomes
```





# Post-CAHOOTS Call Outcome Evolution in Eugene

```{r most frequent natures}

top_14_cahoots_natures <- c(
  "Public Assist",
  "Check Welfare",
  "Transport",
  "Suicidal Subject",
  "Intoxicated Subject",
  "Disorderly Subject",
  "Found Syringe",
  "Info / Atl",
  "Disoriented Subject",
  "Dispute",
  "Subject Down",
  "Disorderly Juveniles",
  "Suspicious Conditions",
  "Police Officer Hold"
)

top_mcslc_natures <- c(
  "Agitation Or Disruptive Behavior",
  "Disorganized Behavior",
  "Difficulty Functioning",
  "Needing Social/Mental Health Services",
  "Suicidal Subject",
  "Harm/Risk Of Harm To Self/Others/Property",
  "Adult Social/Interpersonal Problems",
  "Paranoia",
  "Substance Use",
  "Other",
  "Concerns About Treatment Engagement",
  "Trauma",
  "Running Away",
  "Seeking Mental Health Services"
)

```

## Arrests (all agencies)

Now moving on the the call outcome evolution analysis, we start by harmonizing the data. We calculate the weekly rolling average arrest rate per week for Eugene and Springfield, using only CAHOOTS call types regardless of the agency in charge of the intervention.

```{r filtered arrest weekly data}
date_mcslc_start <- as.POSIXct("2024-08-18")
date_cahoots_stop <- as.POSIXct("2025-04-07")
date_end <- as.POSIXct("2025-12-17") 

panel_arrest <- data %>%
  filter(
    city %in% c("Eugene", "Springfield"),
    timestamp >= as.POSIXct("2023-01-01"),
    nature %in% top_14_cahoots_natures      
  ) %>%
  mutate(
    week    = floor_date(timestamp, "week"),
    treated = as.integer(city == "Eugene"),
    post    = as.integer(timestamp >= date_cahoots_stop),
    arrest  = as.integer(str_detect(str_to_lower(outcome),
                "arrest|juvenile taken into custody"))
  ) %>%
  group_by(city, week, treated) %>%
  summarise(
    n_calls     = n(),
    n_arrests   = sum(arrest, na.rm = TRUE),
    arrest_rate = n_arrests / n_calls,
    post        = max(post),
    .groups     = "drop"
  ) %>%
  mutate(week_num = as.numeric(difftime(week, min(week), units = "weeks")))

```


### Parallel Trends

#### Line Plot

Here is a plot of the trends side by side for a visual overview. We can see that the arrest rates trend for each city is quite similar before the discontinuation, but Eugene's volume is superior. After CAHOOTS stopped servicing Eugene, we observe a surge in arrest whereas Springfield's stays similar. 

```{r parallel trends}

vline_date <- as.Date("2025-04-07")

parallel_trends_check <- panel_arrest %>%
  # MM(4)
  group_by(city) %>%
  arrange(week) %>%
  mutate(arrest_rate_smooth = zoo::rollmean(arrest_rate, k = 4, fill = NA, align = "right")) %>%
  ggplot(aes(x = week, y = arrest_rate_smooth, color = city)) +
  geom_line(linewidth = 0.9) +
  geom_vline(xintercept = as.POSIXct(vline_date),
             linetype = "dashed", color = "red", linewidth = 0.8) +
  annotate("text", x = as.POSIXct(vline_date), 
           y = Inf, label = "CAHOOTS stops\n(Eugene)", 
           vjust = 1.5, hjust = -0.1, color = "red", size = 3.5) +
  scale_color_manual(values = c("Eugene" = "#2ACCFE", "Springfield" = "#FEA4E6")) +
  scale_y_continuous(labels = scales::percent_format(accuracy = 0.1)) +
  labs(
    title    = "Arrest Rate by City",
    subtitle = "4-week rolling average for the 14 most frequent CAHOOTS call natures",
    x = NULL, y = "Arrest Rate", color = "City"
  ) +
  theme_minimal() 

ggsave("../../figures/analysis/arrest_rate_by_city.png", parallel_trends_check, width = 10, height = 5, dpi = 180)

parallel_trends_check


```


#### Linear Trends
The linear trends shows with more precision that Eugene saw an increase in arrests and Springfield's rate is relatively constant over time.

```{r linear trends}

linear_trends <- panel_arrest %>%
  ggplot(aes(x = week, y = arrest_rate, color = city)) +
  geom_line(alpha = 0.3, linewidth = 0.4) +
  geom_smooth(method = "lm", se = TRUE, linewidth = 1) +
  geom_vline(xintercept = as.POSIXct("2025-04-07"),
             linetype = "dashed", color = "red") +
  facet_wrap(~city, scales = "free_y") +  
  scale_color_manual(values = c("Eugene" = "#2ACCFE", "Springfield" = "#FEA4E6")) +
  scale_y_continuous(labels = scales::percent_format(accuracy = 0.1)) +
  labs(title = "Arrest rate and linear trend by city",
       subtitle = "free y-axis scale per city") +
  theme_minimal()

ggsave("../../figures/analysis/linear_trends.png", linear_trends, width = 10, height = 5, dpi = 180)

linear_trends
```

```{r linear trends }
cutoff <- as.POSIXct("2025-04-07")

linear_trends_split <- panel_arrest %>%
  ggplot(aes(x = week, y = arrest_rate, color = city)) +
  geom_line(alpha = 0.3, linewidth = 0.4) +
  
  geom_smooth(
    data = ~ filter(.x, city == "Eugene", week < cutoff),
    method = "lm", se = TRUE, linewidth = 1
  ) +
  geom_smooth(
    data = ~ filter(.x, city == "Eugene", week >= cutoff),
    method = "lm", se = TRUE, linewidth = 1
  ) +
  
  geom_smooth(
    data = ~ filter(.x, city == "Springfield"),
    method = "lm", se = TRUE, linewidth = 1
  ) +
  
  geom_vline(
    xintercept = as.numeric(cutoff),
    linetype = "dashed", color = "red", linewidth = 0.8
  ) +
  geom_text(
    data = data.frame(
      city  = c("Eugene", "Springfield"),
      label = c("CAHOOTS stops", ""),
      x     = c(cutoff, cutoff)
    ),
    aes(x = x, y = Inf, label = label),
    vjust = 1.5, hjust = -0.1, color = "red", size = 3,
    inherit.aes = FALSE
  ) +
  
  facet_wrap(~city, scales = "free_y") +
  scale_color_manual(
    values = c("Eugene" = "#2ACCFE", "Springfield" = "#FEA4E6"),
    name = "City"
  ) +
  scale_y_continuous(labels = scales::percent_format(accuracy = 0.1)) +
  labs(
    title    = "Arrest rate and linear trend by city",
    subtitle = "Separate pre/post trends for Eugene; single trend for Springfield (control)",
    x = NULL, y = "Arrest Rate"
  ) +
  theme_minimal()

ggsave("../../figures/analysis/linear_trends_split.png", linear_trends_split, 
       width = 10, height = 5, dpi = 180)
linear_trends_split

```


#### Statistical Test 

We then test the parallel trends more rigorously to see if we can use the Differences in Differences method.

H0 : both groups are the same before treatment (parallel trends)
H1 : trends are different

we don't want to reject null hypothesis so p-value should be over 0.5 (for 95% certainty)

For the overall temporal trend in arrest rate for both cities, the p-value is 0.443 > 0.05, so not significant. It means that there is no temporal drift prior to the treatment period. The baseline for both cities was stable with no significant upward or downward trend common to both cities.

The p-value for `treated`, which is the level difference between Eugene and Springfield, is significant (<0.01). 

Here p-value is 0.544 so the parallel trends passed, meaning that before CAHOOTS was discontinued, arrest rate trends were similar in Eugene and Springfield for call natures typically handled by CAHOOTS, regardless of the agency in charge. We can proceed with the DiD method.

```{r stat test parallel trends}
pre_cahoots <- panel_arrest %>% filter(post == 0)

trend_test_cahoots <- lm(
  arrest_rate ~ week_num * treated,
  data    = pre_cahoots,
  weights = n_calls
)

summary(trend_test_cahoots)
summary(trend_test_cahoots)$coefficients["week_num:treated", c("t value", "Pr(>|t|)")]

stargazer(trend_test_cahoots, type = "text")

stargazer(trend_test_cahoots,
          type = "text",
          title = "Parallel Trends Check",
          covariate.labels = c("Week Number",
                               "Treated",
                               "Week Number:Treated"),
          dep.var.labels = "Arrest Rate")
```

### Differences in Differences

We test the changes in arrest rate using the DiD method. We find (95% confidence level) an increase of 3.2 percentage points for weekly arrest rate in Eugene compared to Springfield for same period after treatment (CAHOOTS stop)


```{r}
did <- lm(
  arrest_rate ~ treated * post,
  data    = panel_arrest,
  weights = n_calls
)

stargazer(did,
          type  = "text",
          title = "DiD, Effect of CAHOOTS Discontinuation on Arrest Rate (Weekly)",
          covariate.labels = c("Eugene (Treated)",
                               "Post-Discontinuation",
                               "DiD Estimator (Treated × Post)"),
          dep.var.labels = "Arrest Rate",
          notes = "OLS weighted by weekly call volume.")


```

#### Event Study

If we take a closer look at the coefficients, grouped by bins of 4 weeks, we can calculate the confidence intervals and plot the arrest rate rolling average, the DiD coefficients.

```{r arrest bin DID}

treatment_week_num <- panel_arrest %>%
  filter(post == 1, city == "Eugene") %>%
  pull(week_num) %>% min()

panel_arrest <- panel_arrest %>%
  mutate(
    rel_week = round(week_num - treatment_week_num),
    rel_bin = floor(rel_week / 4) * 4)   

event_study_binned <- lm(
  arrest_rate ~ treated * factor(rel_bin),
  data    = panel_arrest,
  weights = n_calls
)

coefs_binned <- tidy(event_study_binned) %>%
  filter(str_detect(term, "treated:factor")) %>%
  mutate(
    rel_bin   = as.numeric(str_extract(term, "-?[0-9]+")),
    conf_low  = estimate - 1.96 * std.error,
    conf_high = estimate + 1.96 * std.error
  )

```

```{r DiD coeff plot arrests}

event_study_arrest <- ggplot(coefs_binned, aes(x = rel_bin, y = estimate)) +
  geom_ribbon(aes(ymin = conf_low, ymax = conf_high),
              fill = "gray70", alpha = 0.4) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "gray50") +
  geom_vline(xintercept = -2, linetype = "dashed",
             color = "red", linewidth = 0.8) +
  geom_line(aes(color = rel_bin >= 0), linewidth = 0.8) +
  geom_point(aes(color = rel_bin >= 0), size = 2) +
  geom_smooth(data = filter(coefs_binned, rel_bin >= 0),
              aes(x = rel_bin, y = estimate),
              method = "loess", se = FALSE,
              color = "#C0392B", linewidth = 1.2, linetype = "solid") +
  geom_smooth(data = filter(coefs_binned, rel_bin < 0),
              aes(x = rel_bin, y = estimate),
              method = "loess", se = FALSE,
              color = "#1A8FBB", linewidth = 1.2, linetype = "solid") +
  scale_color_manual(
    values = c("FALSE" = "#2ACCFE", "TRUE" = "#E63946"),
    labels = c("Pre-treatment", "Post-treatment"),
    name   = NULL
  ) +
  scale_y_continuous(labels = scales::percent_format(accuracy = 0.1)) +
  annotate("text", x = 2, y = min(coefs_binned$conf_low, na.rm = TRUE),
           label = "CAHOOTS stops", hjust = 0, color = "red", size = 3) +
  labs(
    title    = "CAHOOTS' discontinuation effect on arrest rate in Eugene",
    subtitle = "4-week bins DiD coefficients relative to bin −1 (reference period)",
    x        = "4-week bins relative to CAHOOTS discontinuation (April 2025)",
    y        = "Effect on arrest rate",
    caption  = "Ref = bin −1. Shaded area = 95% CI. Weighted by call volume."
  ) +
  theme_minimal() +
  theme(legend.position = "top")

ggsave("../../figures/analysis/event_study_arrest.png", event_study_arrest, width = 10, height = 5, dpi = 180)

event_study_arrest

```


#### Robustness Test Arrests

Try different time window to see if it impacts previous results (it passed)

```{r}
panel_narrow <- panel_arrest %>%
  filter(abs(rel_week) <= 12)

did_narrow <- lm(
  arrest_rate ~ treated * post,
  data    = panel_narrow,
  weights = n_calls
)

stargazer(did, did_narrow,
          type = "text",
          column.labels = c("Full period (2023-2025)", 
                            "Narrow window (±12 months)"),
          covariate.labels = c("Eugene (Treated)",
                               "Post-Discontinuation",
                               "DiD Estimator (Treated × Post)"),
          dep.var.labels = "Arrest Rate",
          notes = "Robustness check: restricting to ±12 months around discontinuation.")
```

## Police call outcome

We check for a more detailed analysis to see how police call outcome changed after CAHOOTS stopped servicing Eugene to answer the question:

did the rate of specific outcome change in Eugene relative to Springfield for law enforcement after April 7, 2025

```{r filter EPD}

outcomes_epd <- c(
  "Arrest",
  "Cited In Lieu Of Custody",
  "Juvenile Taken Into Custody",
  "Advised",
  "Resolved",
  "Welfare Check Done",
  "Unable To Locate",
  "Gone On Arrival",
  "Non Criminal Hold",
  "Transport Made",
  "Referred To Other Agency"
)

panel_epd_outcomes <- data %>%
  filter(
    city %in% c("Eugene", "Springfield"),
    timestamp >= as.POSIXct("2023-01-01"),
    nature %in% top_14_cahoots_natures,
    EPD == 1 | SPD == 1
  ) %>%
  mutate(
    week    = floor_date(timestamp, "week"),
    treated = as.integer(city == "Eugene"),
    post    = as.integer(timestamp >= date_cahoots_stop),
    # Create binary columns here in mutate, where outcome IS available
    arrest                    = as.integer(outcome == "Arrest"),
    cited_in_lieu_of_custody  = as.integer(outcome == "Cited In Lieu Of Custody"),
    juvenile_taken_into_custody = as.integer(outcome == "Juvenile Taken Into Custody"),
    advised                   = as.integer(outcome == "Advised"),
    resolved                  = as.integer(outcome == "Resolved"),
    welfare_check_done        = as.integer(outcome == "Welfare Check Done"),
    unable_to_locate          = as.integer(outcome == "Unable To Locate"),
    gone_on_arrival           = as.integer(outcome == "Gone On Arrival"),
    non_criminal_hold         = as.integer(outcome == "Non Criminal Hold"),
    transport_made            = as.integer(outcome == "Transport Made"),
    referred_to_other_agency  = as.integer(outcome == "Referred To Other Agency")
  ) %>%
  group_by(city, week, treated) %>%
  summarise(
    n_calls                       = n(),
    post                          = max(post),
    arrest_rate                   = sum(arrest)                     / n(),
    cited_in_lieu_of_custody_rate = sum(cited_in_lieu_of_custody)   / n(),
    juvenile_taken_into_custody_rate = sum(juvenile_taken_into_custody) / n(),
    advised_rate                  = sum(advised)                    / n(),
    resolved_rate                 = sum(resolved)                   / n(),
    welfare_check_done_rate       = sum(welfare_check_done)         / n(),
    unable_to_locate_rate         = sum(unable_to_locate)           / n(),
    gone_on_arrival_rate          = sum(gone_on_arrival)            / n(),
    non_criminal_hold_rate        = sum(non_criminal_hold)          / n(),
    transport_made_rate           = sum(transport_made)             / n(),
    referred_to_other_agency_rate = sum(referred_to_other_agency)   / n(),
    .groups = "drop"
  ) %>%
  mutate(week_num = as.numeric(difftime(week, min(week), units = "weeks")))

```

Function to iterate through the call outcome list and do parallel trends test, did if possible, summary if not

```{r did iterate}
run_did_single <- function(outcome_rate_col, panel, label) {
  
  formula_trend <- reformulate(c("week_num * treated"), response = outcome_rate_col)
  formula_did   <- reformulate(c("treated * post"),     response = outcome_rate_col)
  
  pre  <- panel %>% filter(post == 0)
  trend <- lm(formula_trend, data = pre, weights = n_calls)
  trend_p <- summary(trend)$coefficients["week_num:treated", "Pr(>|t|)"]
  
  cat("\n***************************************\n")
  cat("\n***************************************\n")
  cat("OUTCOME:", label, "\n")
  cat("Parallel trends p-value (week_num:treated):", round(trend_p, 4), "\n")
  
  if (trend_p > 0.05) {
    cat("Parallel trends hold -> proceeding with DiD\n")
    did <- lm(formula_did, data = panel, weights = n_calls)
    stargazer(did,
              type  = "text",
              title = paste("DiD —", label),
              covariate.labels = c("Eugene (Treated)", 
                                   "Post-Discontinuation",
                                   "DiD Estimator (Treated × Post)"),
              dep.var.labels = label)
  } else {
    cat("Parallel trends FAILED -> DiD not valid, reporting descriptive only\n")
    data %>%
      filter(
        city == "Eugene",
        timestamp >= as.POSIXct("2023-01-01"),
        nature %in% top_14_cahoots_natures,
        EPD == 1
      ) %>%
      mutate(
        post_label = if_else(timestamp >= date_cahoots_stop, 
                             "Post-CAHOOTS", "Pre-CAHOOTS"),
        is_outcome = as.integer(outcome == label)
      ) %>%
      group_by(post_label) %>%
      summarise(
        n_calls  = n(),
        n_outcome = sum(is_outcome),
        rate      = n_outcome / n_calls
      ) %>%
      print()
  }
}


```


Parallel trends test for each relevant call outcome. 


```{r run did test, warning = FALSE}
outcome_cols <- str_to_lower(str_replace_all(outcomes_epd, " ", "_"))

walk2(
  paste0(outcome_cols, "_rate"),
  outcomes_epd,
  ~ run_did_single(.x, panel_epd_outcomes, .y)
)
```



## Call Type Breakdown

### Descriptive

```{r}
arrest_by_nature_eugene <- data %>%
  filter(
    city == "Eugene",
    timestamp >= as.POSIXct("2023-01-01"),
    nature %in% top_14_cahoots_natures,
    EPD == 1
  ) %>%
  mutate(
    post = if_else(timestamp >= date_cahoots_stop, 
                   "Post-CAHOOTS", "Pre-CAHOOTS"),
    arrest = as.integer(str_detect(
      str_to_lower(as.character(outcome)),
      "arrest|juvenile taken into custody"
    ))
  ) %>%
  group_by(nature, post) %>%
  summarise(
    n_calls   = n(),
    n_arrests = sum(arrest, na.rm = TRUE),
    rate      = n_arrests / n_calls,
    .groups   = "drop"
  ) %>%
  pivot_wider(
    names_from  = post,
    values_from = c(n_calls, n_arrests, rate)
  ) %>%
  mutate(
    delta_pp       = `rate_Post-CAHOOTS` - `rate_Pre-CAHOOTS`,
    relative_change = delta_pp / `rate_Pre-CAHOOTS`
  ) %>%
  arrange(desc(delta_pp))

arrest_by_nature_eugene %>%
  knitr::kable(
    digits  = 4,
    caption = "Arrest rate change by call nature — Eugene EPD, before vs after CAHOOTS discontinuation"
  )
```


```{r}
ggplot(arrest_by_nature_eugene, 
       aes(x = reorder(nature, delta_pp), 
           y = delta_pp, 
           fill = delta_pp > 0)) +
  geom_col(width = 0.7) +
  geom_text(aes(label = scales::percent(delta_pp, accuracy = 0.1)),
            hjust = ifelse(arrest_by_nature_eugene$delta_pp > 0, -0.1, 1.1),
            size = 3) +
  coord_flip() +
  scale_fill_manual(values = c("TRUE" = "#E63946", "FALSE" = "#2ACCFE"),
                    labels = c("Decrease", "Increase"),
                    name   = "Direction") +
  scale_y_continuous(labels = scales::percent_format(accuracy = 0.1)) +
  labs(
    title    = "Change in arrest rate by call nature after CAHOOTS discontinuation",
    subtitle = "Eugene EPD only, descriptive",
    x        = NULL,
    y        = "Change in arrest rate (pp)"
  ) +
  theme_minimal()
```



### DiD Springfield comparison


```{r}
run_did_by_nature <- function(nature_filter, panel_data) {
  
  df <- panel_data %>%
    filter(nature == nature_filter)
  
  # Need enough observations
  if (nrow(df) < 50) return(NULL)
  
  # Parallel trends check
  pre <- df %>% filter(post == 0)
  
  if (nrow(pre) < 20) return(NULL)
  
  trend <- lm(arrest_rate ~ week_num * treated, 
              data = pre, weights = n_calls)
  trend_p <- tryCatch(
    summary(trend)$coefficients["week_num:treated", "Pr(>|t|)"],
    error = function(e) NA
  )
  
  # DiD
  did <- lm(arrest_rate ~ treated * post,
            data = df, weights = n_calls)
  
  did_coef <- tidy(did) %>%
    filter(term == "treated:post") %>%
    select(estimate, std.error, p.value)
  
  tibble(
    nature          = nature_filter,
    parallel_trends = trend_p,
    pt_passed       = !is.na(trend_p) & trend_p > 0.05,
    did_estimate    = did_coef$estimate,
    did_se          = did_coef$std.error,
    did_p           = did_coef$p.value,
    significant     = !is.na(did_coef$p.value) & did_coef$p.value < 0.05
  )
}

# Build nature-level panel first
panel_by_nature <- data %>%
  filter(
    city %in% c("Eugene", "Springfield"),
    timestamp >= as.POSIXct("2023-01-01"),
    nature %in% top_14_cahoots_natures,
    EPD == 1 | SPD == 1
  ) %>%
  mutate(
    week    = floor_date(timestamp, "week"),
    treated = as.integer(city == "Eugene"),
    post    = as.integer(timestamp >= date_cahoots_stop),
    arrest  = as.integer(str_detect(
      str_to_lower(as.character(outcome)),
      "arrest|juvenile taken into custody"
    ))
  ) %>%
  group_by(city, nature, week, treated) %>%
  summarise(
    n_calls     = n(),
    n_arrests   = sum(arrest, na.rm = TRUE),
    arrest_rate = n_arrests / n_calls,
    post        = max(post),
    .groups     = "drop"
  ) %>%
  mutate(week_num = as.numeric(difftime(week, min(week), units = "weeks")))

# Run for all natures
results_by_nature <- map_dfr(
  top_14_cahoots_natures,
  ~ run_did_by_nature(.x, panel_by_nature)
) %>%
  arrange(desc(abs(did_estimate)))

results_by_nature %>%
  knitr::kable(
    digits  = 4,
    caption = "Nature-stratified DiD — arrest rate change per call type"
  )
```


```{r}
results_by_nature %>%
  filter(pt_passed) %>%
  mutate(
    conf_low  = did_estimate - 1.96 * did_se,
    conf_high = did_estimate + 1.96 * did_se,
    color     = case_when(
      significant & did_estimate > 0 ~ "Significant increase",
      significant & did_estimate < 0 ~ "Significant decrease",
      TRUE ~ "Not significant"
    )
  ) %>%
  ggplot(aes(x = reorder(nature, did_estimate), 
             y = did_estimate, color = color)) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "gray50") +
  geom_errorbar(aes(ymin = conf_low, ymax = conf_high), 
                width = 0.3, linewidth = 0.7) +
  geom_point(size = 3) +
  coord_flip() +
  scale_color_manual(values = c(
    "Significant increase" = "#E63946",
    "Significant decrease" = "#2ACCFE",
    "Not significant"      = "gray60"
  )) +
  scale_y_continuous(labels = scales::percent_format(accuracy = 0.1)) +
  labs(
    title    = "DiD estimates by call nature, effect on arrest rate",
    subtitle = "Only natures where parallel trends hold | 95% CI shown",
    x        = NULL,
    y        = "DiD estimate (pp change in arrest rate)",
    color    = NULL
  ) +
  theme_minimal() +
  theme(legend.position = "top")
```





### DiD interaction model

```{r}
did_interaction <- lm(
  arrest_rate ~ treated * post * nature,
  data    = panel_by_nature,
  weights = n_calls
)

stargazer(did_interaction, type="text")

# Extract nature-specific DiD coefficients
tidy(did_interaction) %>%
  filter(str_detect(term, "treated:post:nature")) %>%
  mutate(
    nature    = str_remove(term, "treated:post:nature"),
    conf_low  = estimate - 1.96 * std.error,
    conf_high = estimate + 1.96 * std.error
  ) %>%
  arrange(desc(estimate)) %>%
  knitr::kable(digits = 4)


```


```{r}
# Extract base DiD (reference nature = Check Welfare)
base_did <- tidy(did_interaction) %>%
  filter(term == "treated:post") %>%
  pull(estimate)

# Extract nature-specific deviations and compute total effect
nature_did_total <- tidy(did_interaction) %>%
  filter(str_detect(term, "treated:post:nature")) %>%
  mutate(
    nature        = str_remove(term, "treated:post:nature"),
    total_did     = estimate + base_did,
    conf_low      = total_did - 1.96 * std.error,
    conf_high     = total_did + 1.96 * std.error
  ) %>%
  # Add reference nature manually
  bind_rows(
    tidy(did_interaction) %>%
      filter(term == "treated:post") %>%
      mutate(
        nature    = "Check Welfare",
        total_did = estimate,
        conf_low  = estimate - 1.96 * std.error,
        conf_high = estimate + 1.96 * std.error
      )
  ) %>%
  arrange(desc(total_did))

nature_did_total %>%
  select(nature, total_did, std.error, p.value, conf_low, conf_high) %>%
  knitr::kable(
    digits  = 4,
    caption = "Total nature-specific DiD estimates from pooled interaction model"
  )
```


```{r}
# Identify natures with enough observations in all 4 cells
# (Eugene/Springfield x Pre/Post)
valid_natures <- panel_by_nature %>%
  group_by(nature, treated, post) %>%
  summarise(n = sum(n_calls), .groups = "drop") %>%
  group_by(nature) %>%
  summarise(
    n_cells    = n(),
    min_calls  = min(n),
    .groups    = "drop"
  ) %>%
  filter(n_cells == 4, min_calls >= 10) %>%  # all 4 cells exist with enough data
  pull(nature)

cat("Valid natures for interaction model:", length(valid_natures), "\n")
print(valid_natures)

# Refit on valid natures only
panel_by_nature_valid <- panel_by_nature %>%
  filter(nature %in% valid_natures)

did_interaction_clean <- lm(
  arrest_rate ~ treated * post * nature,
  data    = panel_by_nature_valid,
  weights = n_calls
)

# Now the F-test should work
valid_terms_clean <- grep("treated:post:nature",
                          names(coef(did_interaction_clean)),
                          value = TRUE)
valid_terms_clean <- valid_terms_clean[
  !is.na(coef(did_interaction_clean)[valid_terms_clean])
]

linearHypothesis(did_interaction_clean, valid_terms_clean)

```













## All calls

we answered the question "within CAHOOTS call types, do specific natures drive the arrest increase more than others in Eugene?" using Springfield as a control group.

Now we take a look at every call to see if the arrest rate increase is specifically concentrated in CAHOOTS call types. Springfield is not our comparison point, we use other call types to compare.






# TODO

TO DO THURSDAY

- put graph in report
- explain method in report
- add theory behind method and statistical testing (equations etc)

- do DiD but for other outcomes
- write State of the Art
- wrap up 

TO DO FRIDAY 

- check calls from police only and see changes in outcomes 
- decompose call type to see which calls have more arrests
- is police going to more calls; agency mismatch?
- checck again parallel trends


# Appendix

## Abbreviations

- SPD = Springfield Police Department
- CAD = Computer-Aided Dispatch
- MCS LC = Mobile Crisis Services Lane County







- EPD: Eugene Police Department
- CAHOOTS: Crisis Assistance Helping Out On the Streets
- MCSLC : Mobile Crisis Services Lane County
- SPD: Springfield Police Department

- MCR: Mobile Crisis Response









