---
title: "Non-violent emergency call outcome analysis in Eugene, Oregon"
subtitle: "Investigating Eugene's response to addiction, homelessness and mental health related crisis"
author: "Mathilde Lafay"
date: "`r Sys.Date()`"
output:
  pdf_document:
    toc: true
    toc_depth: 3
    number_sections: true
    fig_caption: true
    df_print: kable
    latex_engine: xelatex
    pandoc_args: ["--syntax-highlighting", "tango"]
include-before:
  - \newpage
---
\newpage
# Introduction 

In this document, we will be conducting a quantitative analysis of the call outcomes for mental health, addiction and homelessness related emergencies in Eugene. We will be focusing on the code and methods. For more details about the project, please find the full report at the root of the repository: `internship_report.pdf`.

# Work Environment

## Prerequisites
- [R](https://cran.r-project.org/) (version 4.5 or higher recommended)
- [RStudio](https://posit.co/download/rstudio-desktop/) (optional, but recommended)

## Managing Dependencies
This project uses [`renv`](https://rstudio.github.io/renv/) for R package management. To install all required packages:

1. Open the project in RStudio or start an R session in the project directory.
2. Run:
	```R
	renv::restore()
	```
	This will install all packages listed in the project's lockfile.

If you add new packages, use `renv::snapshot()` to update the lockfile.

## Libraries

```{r - libraries, echo = TRUE, message = FALSE, warning = FALSE}

library(here) # for relative paths
library(tidyverse)
library(dplyr)

library(readxl)
library(purrr)
library(janitor)
library(lubridate)
library(hms)

```

## Color Palette

```{r - color palette}

# color blind friendly palette https://colors-tester.com/free/fr/
color_agencies <- c(
  "EPD"     = "#2ACCFE",  
  "SPD"     = "#FEA4E6",  
  "CAHOOTS" = "#58BF03",  
  "MCS_LC"   = "#FF8C66"   
)



```

# Data

Here is the command to open the clean data.

```{r - data}

data <- readRDS(here("data/clean/data_merged.rds"))
head(data)

```

We processed it from the following sources.

## Available raw data

- SPD Call for Services, 2015-2025 : .xlsx document, one sheet per year
- SPD Responding Units, 2015-2025 : .xlsx document, one sheet per year
- SPD Call Outcome, 2015-2025 : .xlsx doxument, key codes and one sheet per year

- MCS LC data, 2024/08/18 - 2025/12/17 : .xlsx document, one sheet

- Eugene CAD data, 2015-2025 : one .csv per year

### SPD data

This dataset contains detailed records of calls for service for Springfield Police Department from 2015 to 2025, including temporal management of calls, the nature of the interventions, and the mobile resources deployed.

There are 14 columns and 567,655 rows.

| Variable               | Type      | Example Value         | Description                                                                    |
|------------------------|-----------|-----------------------|--------------------------------------------------------------------------------|
| `timestamp`            | datetime  | 2015-01-01 01:16:21   | Date and time when the incident was initially recorded.                        |
| `incident_id`          | double    | 15000074              | Unique identifier for each incident.                                           |
| `agency`               | factor    | SPD                   | Agency responsible for the intervention.                                       |
| `city`                 | factor    | Springfield           | City.                                                                          |
| `nature_code`          | factor    | ALMAUD                | Coded category for the type of call received.                                  |
| `nature_final`         | factor    | AUDIBLE ALARM         | Final classification of the incident after assessment.                         |
| `close_code`           | factor    | BLDS                  | Coded reason or outcome for closing the incident.                              |
| `close_code_definition`| factor    | Building Check Secure | Full description of the incident closure status.                               |
| `priority`             | factor    | 3                     | Numerical priority level assigned to the call.                                 |
| `prime_unit`           | character | 1S18                  | Primary unit assigned to respond to the incident.                              |
| `units`                | character | 1S18, 1S21, K94       | List of all units dispatched to the scene.                                     |
| `dispatch_time`        | time      | 01:18:00              | Time at which units were dispatched.                                           |
| `arrival_time`         | time      | 01:21:00              | Time at which the first unit arrived at the scene.                             |
| `clear_time`           | time      | 01:32:00              | Time at which the incident was marked as completed.                            |
| `call_source`          | factor    | PHONE                 | Origin or method of the incoming call.                                         |



### Eugene CAD data

This dataset contains detailed records of calls for service for Eugene from 2015 to 2025, including temporal management of calls, the nature of the interventions, and the mobile resources deployed.

There are 17 columns and 1,446,014 rows.

| Variable              | Type      | Example Value         | Description                                                                    |
|-----------------------|-----------|-----------------------|--------------------------------------------------------------------------------|
| `timestamp`           | datetime  | 2015-01-01 00:00:00   | Date and time when the call to report an incident was initially recorded.      |
| `incident_id`         | double    | 15000001              | Unique identifier for each incident.                                           |
| `agency`              | factor    | EPD                   | Agency responsible for the intervention (e.g., Eugene Police Department).      |
| `service`             | factor    | LAW                   | Type of service mobilized (e.g., Law Enforcement, Fire, Other).                |
| `call_source`         | factor    | SELF                  | Origin of the call (e.g., Self-initiated, Phone, W911).                        |
| `nature`              | factor    | PERSON STOP           | Nature or reported motive of the incident.                                     |
| `close_code`          | factor    | ASST                  | Abbreviated closing code indicating the outcome of the intervention.           |
| `closed_as`           | factor    | ASSISTED              | Full description of how the incident was resolved.                             |
| `priority`            | factor    | 6                     | Priority level assigned to the incident ranging from 1 (highest) to 9 (lowest).|
| `dispatch_time`       | time      | 00:00:00              | Time when units were dispatched to the incident location.                      |
| `arrival_time`        | time      | 00:00:00              | Time when the first unit arrived on the scene.                                 |
| `clear_time`          | time      | 00:03:37              | Time when the incident was marked as cleared or completed.                     |
| `dispatch`            | factor    | 1                     | Binary indicator specifying whether a unit was dispatched.                     |
| `arrival`             | factor    | 1                     | Binary indicator specifying whether a unit arrived on the scene.               |
| `prime_unit`          | factor    | _5E48                 | Identifier of the primary unit assigned to the incident.                       | 
| `nb_units_dispatched` | double    | 1                     | Total number of units dispatched to the intervention.                          |
| `nb_units_arrived`    | double    | 1                     | Total number of units that successfully reached the scene.                     |


### Mobile Crisis Services - Lane County data

This dataset tracks crisis intervention activities from MCS LC, focusing on response timelines, and clinical outcomes for mobile crisis units.

| Variable                     | Type      | Example Value         | Description                                                                 |
|------------------------------|-----------|-----------------------|-----------------------------------------------------------------------------|
| `timestamp`                  | datetime  | 2025-12-17 02:57:00   | Initial date and time the crisis call was logged.                           |
| `incident_id`                | double    | 13209                 | Unique identifier for the crisis incident.                                  |
| `agency`                     | factor    | MCS LC                | The specific Mobile Crisis Service agency responding to the call.           |
| `city`                       | factor    | Springfield           | The city or municipality where the incident occurred.                       |
| `call_type`                  | factor    | Agitation             | The primary clinical nature or behavioral reason for the call.              |
| `call_outcome`               | factor    | Remained in community | The final disposition or result of the intervention.                        |
| `dispatch_time`              | datetime  | 2025-12-17 03:14:00   | Time when the mobile crisis team was officially dispatched.                 |
| `client_engagement_time`     | datetime  | 2025-12-17 03:38:00   | Time when the team made direct clinical contact with the client.            |
| `arrival_time`               | datetime  | 2025-12-17 03:35:00   | Time when the mobile crisis unit arrived at the scene.                      |
| `clear_time`                 | datetime  | 2025-12-17 03:48:00   | Time when the unit completed the intervention and became available again.   |
| `dispatch_status`            | factor    | Engaged               | Status of the dispatch (e.g., Engaged, Refused).                            |
| `minutes_request_dispatch`   | double    | 17                    | Duration in minutes between the initial request and dispatch.               |
| `minutes_dispatch_arrival`   | double    | 21                    | Duration in minutes between dispatch and arrival on scene.                  |
| `minutes_arrival_engagement` | double    | 3                     | Duration in minutes between arrival and start of client engagement.     |
| `minutes_arrival_departure`  | double    | 13                    | Total time spent on scene in minutes (arrival to departure).                |




# Analysis

The following analysis will focus on investigating the changes in Eugene's response to non-violent emergencies related to homelessness, addiction and mental health after MCS LC's launch on August 18th 2024 and CAHOOTS' discontinuation on April 7th 2025 in Eugene.

## Overview and exploratory analysis

Let's begin with an generic overview and exploratory analysis of the data.

```{r - glimpse}
glimpse(data)
```

```{r - global call volume, echo = TRUE, message = FALSE, warning = FALSE}

data %>%
  group_by(city) %>%
  summarise(
    Total_calls = n(),
    EPD   = sum(EPD == 1),
    SPD   = sum(SPD == 1),
    CAHOOTS = sum(CAHOOTS == 1),
    MCSLC = sum(MCSLC == 1)
  ) %>%
  knitr::kable(caption = "Global repartition of call volume per city and agency")
```


```{r - cahoots natures, echo = TRUE, fig.align = "center", fig.width = 7, fig.height = 4.5, fig.cap = "Top 10 most frequent call natures for CAHOOTS dispatches"}

data_nature_cahoots <- data %>%
  filter(CAHOOTS == 1) %>%
  count(nature, sort = TRUE) %>%
  top_n(20, n)

ggplot(data_nature_cahoots, aes(x = reorder(nature, n), y = n)) +
  geom_col(fill = "#58BF03", width = 0.7) +
  geom_text(
    aes(label = scales::comma(n)),
    hjust = 1.1,                   
    color = "#1D3557",             
    size = 3,                     
    fontface = "bold"
  ) +
  coord_flip() +
  labs(
    title = "Most frequent natures (CAHOOTS)",
    x = "Nature of incident",
    y = "Number of interventions"
  )

```



```{r - evolution in time, echo = TRUE, fig.align = "center", fig.width = 7.5, fig.height = 4.5, fig.cap = "Monthly evolution of call volume by mcs agency (2015-2025)"}

data_chronologie <- data %>%
  mutate(Mois = floor_date(timestamp, "month")) %>%
  group_by(Mois) %>%
  summarise(
    CAHOOTS = sum(CAHOOTS == 1),
    MCS_LC  = sum(MCSLC == 1)
  ) %>%
  pivot_longer(cols = c(CAHOOTS, MCS_LC), names_to = "Service", values_to = "Volume")

ggplot(data_chronologie, aes(x = Mois, y = Volume, color = Service)) +
  geom_line(size = 0.8) +
  scale_color_manual(values = color_agencies) +  
  labs(
    title = "Timeline for Mobile Crisi Services",
    subtitle = "Monthly intervention volume",
    x = "Chronology (Month)",
    y = "Number of dispatched calls"
  ) +
  theme_minimal() 

```



## Call outcome focus








# Appendix

## Abbreviations

- SPD = Springfield Police Department
- CAD = Computer-Aided Dispatch
- MCS LC = Mobile Crisis Services Lane County









